---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway:

```{r load}
if (!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)
library(nycflights13)
```


# Exploratory Data Analysis

## The Theoretic Underpinning

Exploratory data analysis (EDA) is an iterative cycle:

1) Generate questions about your data.

2) Search for answers by visualising, transforming, and modelling your data.

3) Use what you learn to refine your questions and/or generate new questions.

> "Your goal during EDA is to develop an understanding of your data." - R For Data Science

Two useful questions to ask about your data:

1) What type of variation occurs within my variables?

2) What type of covariation occurs between my variables?

To facilitate these questions, let's define some terms:

* **Variable** - a quantity, quality, or property that you can measure.

* **Value** - the state of a variable when you measure it.

* **Observation** - a set of measurements made under similar conditions.

* **Tabular data** -a set of values, each associated with a variable and an observation.


## Variation

**Variation** is the tendency of the values of a variable to change from measurement to measurement.  Every variable has its own pattern of variation, which can reveal interesting information. The best way to understand that pattern is to visualise the distribution of the variable’s values.


### Visualising Distribution

How you visualise the distribution of a variable will depend on whether the variable is categorical or continuous. A variable is **categorical** if it can only take one of a small set of values. In R, categorical variables are usually saved as factors or character vectors. To examine the distribution of a categorical variable, use a bar chart:

```{r}
ggplot(data = diamonds) +
  
  geom_bar(mapping = aes(x = cut))
```


The height of the bars displays how many observations occurred with each x value. You can compute these values manually with `dplyr::count()`:

```{r}
diamonds %>% 
  
  count(cut)
```

A variable is **continuous** if it can take any of an infinite set of ordered values. To examine the distribution of a continuous variable, use a histogram:

```{r}
ggplot(data = diamonds) +
  
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

You can compute this by hand by combining `dplyr::count()` and `ggplot2::cut_width()`:

```{r}
diamonds %>% 
  
  count(cut_width(carat, 0.5)) #??
```

A **histogram** divides the x-axis into equally spaced bins and then uses the height of a bar to display the number of observations that fall in each bin. You should always explore a variety of binwidths when working with histograms, as different binwidths can reveal different patterns. 

```{r}
smaller <- diamonds %>% 
  
  filter(carat < 3)
  
ggplot(data = smaller, mapping = aes(x = carat)) +
  
  geom_histogram(binwidth = 0.1)
```

If you wish to overlay multiple histograms in the same plot, use `geom_freqpoly()`instead of `geom_histogram()`. 

```{r}
ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  
  geom_freqpoly(binwidth = 0.1)
```

Having visualised variation, we need to interrogate the data. Here follow some questions we could ask. 

### Typical Values

Data visulisation helps us to turn descriptive information and typical values into useful questions. Consider the following for the `diamonds` data set:

```{r}
ggplot(data = smaller, mapping = aes(x = carat)) +
  
  geom_histogram(binwidth = 0.01)
```

- Which values are the most common? Why?

- Which values are rare? Why? Does that match your expectations?

- Can you see any unusual patterns? What might explain them?

- Why are there more diamonds at whole carats and common fractions of carats?

- Why are there more diamonds slightly to the right of each peak than there are slightly to the left of each peak?

- Why are there no diamonds bigger than 3 carats?

Clusters of similar values suggest that subgroups exist in your data. To understand the subgroups, ask:

- How are the observations within each cluster similar to each other?

- How are the observations in separate clusters different from each other?

- How can you explain or describe the clusters?

- Why might the appearance of clusters be misleading?

Clusters would look something like the Old Faithful data dataset:

```{r}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  
  geom_histogram(binwidth = 0.25)
```


### Unusual Values

Outliers are observations that are unusual; data points that don’t seem to fit the pattern. Outliers are sometimes hard to spot:

```{r}
ggplot(diamonds) + 
  
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```

To make it easy to see the unusual values, we need to zoom to small values of the y-axis with `coord_cartesian()`.

```{r}
ggplot(diamonds) + 
  
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  
  coord_cartesian(ylim = c(0, 50))
```

`xlim()` and `ylim()` zoom in on the data, or throw away the data outside the limits, depending on how it they are used.

We can see what's going on in the data with some `dplyr` magic:

```{r}
unusual <- 
  
  diamonds %>% 
  
  filter(y < 3 | y > 20) %>% 
  
  select(price, x, y, z) %>%
  
  arrange(y)

unusual
```


It’s good practice to repeat your analysis with and without the outliers. If they have minimal effect on the results, and you can’t figure out why they’re there, it’s reasonable to replace them with missing values, and move on. However, if they have a substantial effect on your results, you shouldn’t drop them without justification. You’ll need to figure out what caused them (e.g. a data entry error) and disclose that you removed them in your write-up.


### Missing Values

Unusual values can be removed from the data set, but this may adversely affect your dataset if you have a low quality data set. Instead, it might be useful to replace your unusual values with `NA`s:

```{r}
diamonds2 <- diamonds %>% 
  
  mutate(y = ifelse(y < 3 | y > 20, NA, y))
```


Use `?ifelse()` to find out more about that function. If you need to mutate based on more (or more complex) criteria, `dplyr::case_when()` comes in really useful. `case_when()` is particularly useful inside mutate when you want to create a new variable that relies on a complex combination of existing variables.

Package ggplot2 will tell you will tell you when it removes missing values:

```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  
  geom_point()
```


Sometimes missing values themselves convey information, and should not just be disregarded as fillers. You might want to understand what makes observations with missing values different to observations with recorded values. For example, in `nycflights13::flights`, missing values in the `dep_time` variable indicate that the flight was cancelled. So you might want to compare the scheduled departure times for cancelled and non-cancelled times. You can do this by making a new variable with `is.na()`.

```{r}
nycflights13::flights %>% 
  
  mutate(
    
    cancelled = is.na(dep_time),
    
    sched_hour = sched_dep_time %/% 100,
    
    sched_min = sched_dep_time %% 100,
    
    sched_dep_time = sched_hour + sched_min / 60
    
  ) %>% 
  
  ggplot(mapping = aes(sched_dep_time)) + 
  
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
```

This comparison can still be improved, though.


### Covariation

If variation describes the behavior **within** a variable, covariation describes the behavior **between** variables. It is the tendency for the values of two or more variables to vary together in a related way.Determining the degree of covariation depends on the types of the variables involved.

If your dataset contained both categorical and continuous variables, for example, it might be desirable to breakdown the continuous variable according to the categorical variable, Here is an example with the `diamonds` data set: 

```{r}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

It is rather hard to see the difference in distribution because the overall counts differ so much:

```{r}
ggplot(diamonds) + 
  
  geom_bar(mapping = aes(x = cut))
```

We can fix this by having `geom_freqpoly()` plot according to density instead of the count. The *density* hereis the count standardised so that the area under each frequency polygon is one. 

```{r}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```














