---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```


<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway:

```{r load}

if (!require("tidyverse")) install.packages("tidyverse")
library("tidyverse")

```


# Exploratory Data Analysis

Exploratory data analysis (EDA) is an iterative cycle:

1) Generate questions about your data.

2) Search for answers by visualising, transforming, and modelling your data.

3) Use what you learn to refine your questions and/or generate new questions.

> "Your goal during EDA is to develop an understanding of your data." - R For Data Science


