---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```


<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway:

```{r load}

if (!require("tidyverse")) install.packages("tidyverse")
library("tidyverse")

```


# Exploratory Data Analysis

## The Theoretic Underpinning

Exploratory data analysis (EDA) is an iterative cycle:

1) Generate questions about your data.

2) Search for answers by visualising, transforming, and modelling your data.

3) Use what you learn to refine your questions and/or generate new questions.

> "Your goal during EDA is to develop an understanding of your data." - R For Data Science

Two useful questions to ask about your data:

1) What type of variation occurs within my variables?

2) What type of covariation occurs between my variables?

To facilitate these questions, let's define some terms:

* **Variable** - a quantity, quality, or property that you can measure.

* **Value** - the state of a variable when you measure it.

* **Observation** - a set of measurements made under similar conditions.

* **Tabular data** -a set of values, each associated with a variable and an observation.


## Variation

**Variation** is the tendency of the values of a variable to change from measurement to measurement.  Every variable has its own pattern of variation, which can reveal interesting information. The best way to understand that pattern is to visualise the distribution of the variable’s values.


### Visualising Distribution

How you visualise the distribution of a variable will depend on whether the variable is categorical or continuous. A variable is **categorical** if it can only take one of a small set of values. In R, categorical variables are usually saved as factors or character vectors. To examine the distribution of a categorical variable, use a bar chart:

```{r}
ggplot(data = diamonds) +
  
  geom_bar(mapping = aes(x = cut))
```


The height of the bars displays how many observations occurred with each x value. You can compute these values manually with `dplyr::count()`:

```{r}
diamonds %>% 
  
  count(cut)
```

A variable is **continuous** if it can take any of an infinite set of ordered values. To examine the distribution of a continuous variable, use a histogram:

```{r}
ggplot(data = diamonds) +
  
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

You can compute this by hand by combining `dplyr::count()` and `ggplot2::cut_width()`:

```{r}
diamonds %>% 
  
  count(cut_width(carat, 0.5)) #??
```

A **histogram** divides the x-axis into equally spaced bins and then uses the height of a bar to display the number of observations that fall in each bin. You should always explore a variety of binwidths when working with histograms, as different binwidths can reveal different patterns. 

```{r}
smaller <- diamonds %>% 
  
  filter(carat < 3)
  
ggplot(data = smaller, mapping = aes(x = carat)) +
  
  geom_histogram(binwidth = 0.1)
```

If you wish to overlay multiple histograms in the same plot, use `geom_freqpoly()`instead of `geom_histogram()`. 

```{r}
ggplot(data = smaller, mapping = aes(x = carat, colour = cut)) +
  
  geom_freqpoly(binwidth = 0.1)
```

Having visualised variation, we need to interrogate the data. Here follow some questions we could ask. 

### Typical Values

  