---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway:

```{r load}
if (!require("tidyverse")) install.packages("tidyverse")
```

# Tibbles

## Creating Tibbles

Tibbles **are** data frames, but they tweak some older behaviours to make life a little easier. Tibbles are one of the unifying features of the tidyverse. Most other R packages use regular data frames, so you might want to coerce a data frame to a tibble. You can do that with `as_tibble()`:

```{r}
as_tibble(iris)
```

You can create a new tibble from individual vectors with `tibble()`, which automatically recycles inputs of length 1. It’s possible for a tibble to have column names that are not valid R variable names, aka *non-syntactic* names. To refer to these variables, you need to surround them with backticks, \\``\\`:

```{r}
tb <- 
  tibble(
    
  `:)` = "smile", 
  
  ` ` = "space",
  
  `2000` = "number"
)

tb
```


Another function, `tribble()`, is customised for data entry in code: column headings are defined by formulas (i.e. they start with ~), and entries are separated by commas.

```{r}
tribble(
  
  ~x, ~y, ~z,
  
  #--|--|----
  
  "a", 2, 3.6,
  
  "b", 1, 8.5
  
)
```


## Tibbles vs. data.frame

There are two main differences in the usage of a tibble vs. a classic `data.frame`: printing and subsetting.


### Printing

Tibbles have a refined print method that shows only the first 10 rows, and all the columns that fit on screen.

```{r}
tibble(
  
  a = lubridate::now() + runif(1e3) * 86400,
  
  b = lubridate::today() + runif(1e3) * 30,
  
  c = 1:1e3,
  
  d = runif(1e3),
  
  e = sample(letters, 1e3, replace = TRUE)
  
)
```


Tibbles are designed so that you don’t accidentally overwhelm your console when you print large data frames. 

You can control the default print behaviour by setting options:

+ `options(tibble.print_max = n, tibble.print_min = m)`: if more than `n` rows, print only `m` rows. 

+ `options(tibble.print_min = Inf)` always shows all rows.

+ `options(tibble.width = Inf)` always prints all columns, regardless of the width of the screen.

+ For a complete list of options, view documentation using `package?tibble`.

+ R-studio's `View()` function is also useful for viewing at the en of long manipulations. 


### Subsetting 

`$` and `[[` can be used to pull out a single variable from a data frame. `[[` extracts by name and value while `$` only extracts by name but is a little less typing.

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
```

<br>

```{r}
# Extract by name
df[["x"]]
```

<br>

```{r}
# Extract by position
df[[1]]
```

To use these in a pipe, you’ll need to use the special placeholder `.`:
```{r}
df %>% .$x
```

```{r}
df %>% .[["x"]]
```

Compared to a `data.frame`, tibbles are more strict: they never do partial matching, and they will generate a warning if the column you are trying to access does not exist.

