---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway:

```{r load}
if (!require("tidyverse")) install.packages("tidyverse")
```

# Data Import

## Loading with `readr`

Most of readr’s functions are concerned with turning flat files into data frames:

+ `read_csv()` - comma delimited files

+ `read_csv2()` - semicolon separated files

+ `read_tsv()` - tab delimited files

+ `read_delim()` - any delimiter

+ `read_fwf()` - fixed width files. You can specify fields either by their widths with `fwf_widths()` or their position with `fwf_positions()`.

+ `read_table()` reads a common variation of fixed width files where columns are separated by white space.

+ `read_log()` reads Apache style log files

When using these functions the nice feature about them is that they print out a column specification that gives the name and type of each column.


## Use Case for `read_csv`

`read_csv()` uses the first line of the data for the column names, which is a very common convention. There are two cases where you might want to tweak this behaviour:

1. Sometimes there are a few lines of metadata at the top of the file. You can use `skip = n` to skip the first `n` lines; or use `comment = "#"` to drop all lines that start with (e.g.) `#`.

```{r}
read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2)
```


```{r}
read_csv("# A comment I want to skip
  x,y,z
  1,2,3", comment = "#")
```


2. The data might not have column names. You can use `col_names = FALSE` to tell read_csv() not to treat the first row as headings.

```{r}
read_csv("1,2,3\n4,5,6", col_names = FALSE)
```


## Compared to base R

If you’ve used R before, you might wonder why we’re not using `read.csv()`. 

* Readr functions are typically much faster (~10x) than their base equivalents. Long running jobs have a progress bar, so you can see what’s happening. If you’re looking for raw speed, try `data.table::fread()`.

* They produce tibbles, they don’t convert character vectors to factors, use row names, or munge the column names. These are common sources of frustration with the base R functions.

* They are more reproducible. Base R functions inherit some behaviour from your operating system and environment variables, so import code that works on your computer might not work on someone else’s.
