---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway:

```{r load}
if (!require("tidyverse")) install.packages("tidyverse")
```

To work with relational data you need verbs that work with pairs of tables. There are three families of verbs designed to work with relational data:

 + _Mutating joins_, which add new variables to one data frame from matching observations in another.

 + _Filtering joins_, which filter observations from one data frame based on whether or not they match an observation in the other table.

 + _Set operations_, which treat observations as if they were set elements.
