---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway:

```{r load}
if (!require("tidyverse")) install.packages("tidyverse")
```

The focus of this chapter will be on regular expressions, or regexps for short. Regular expressions are useful because strings usually contain unstructured or semi-structured data, and regexps are a concise language for describing patterns in strings.This chapter will focus on the stringr package for string manipulation, which is part of the core tidyverse.

# String Basics

You can create strings with either single quotes or double quotes:

```{r}
string1 <- "This is a string"

string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
```

To include a literal single or double quote in a string you can use \ to “escape” it:

```{r}
double_quote <- "\"" # or '"'

single_quote <- '\'' # or "'"
```

Beware that the printed representation of a string is not the same as string itself, because the printed representation shows the escapes. To see the raw contents of the string, use `writeLines()`:

```{r}
x <- c("\"", "\\")

writeLines(x)
```

The most common are "\n", newline, and "\t", tab. You can see the complete list by requesting help on ": ?'"', or ?"'". You’ll also sometimes see strings like "\u00b5", this is a way of writing non-English characters that works on all platforms:

```{r}
x <- "\u00b5"
x
```


Multiple strings are often stored in a character vector, which you can create with `c()`:

```{r}
c("one", "two", "three")
```


## String Length

We'll use some functions from the stringr package. These have more intuitive names, and all start with str_. For example, str_length() tells you the number of characters in a string:

```{r}
str_length(c("a", "R for data science", NA))
```


## Combining Strings

To combine two or more strings, `use str_c()`:

```{r}
str_c("x", "y")

str_c("x", "y", "z")

str_c("x", "y", sep = ", ")
```

If you want to print missing values as "NA", use `str_replace_na()`.

```{r}
x <- c("abc", NA)

str_c("|-", x, "-|")

str_c("|-", str_replace_na(x), "-|")
```

As shown above, `str_c()` is vectorised, and it automatically recycles shorter vectors to the same length as the longest:

```{r}
str_c("prefix-", c("a", "b", "c"), "-suffix")
```

Objects of length 0 are silently dropped. This is particularly useful in conjunction with `if`:

```{r}
name <- "Hadley"

time_of_day <- "morning"

birthday <- FALSE


str_c(
  
  "Good ", time_of_day, " ", name,
  
  if (birthday) " and HAPPY BIRTHDAY",
  
  "."
  
)
```

To collapse a vector of strings into a single string, use `collapse`:

```{r}
str_c(c("x", "y", "z"), collapse = ", ")
```

## Subsetting Strings

You can extract parts of a string using `str_sub()`:

```{r}
x <- c("Apple", "Banana", "Pear")

str_sub(string = x, start = 1, end = 3)
```

Note that `str_sub()` won’t fail if the string is too short; it will just return as much as possible:

```{r}
str_sub("a", 1, 5)
```

You can also use the assignment form of `str_sub()` to modify strings:

```{r}
str_sub(x, 1, 1) <- str_to_lower(str_sub(x, 1, 1))

x
```

## Locales

You can also use `str_to_upper()` or `str_to_title()`. However, changing case is more complicated than it might at first appear because different languages have different rules for changing case. You can pick which set of rules to use by specifying a locale:

```{r}
# Turkish has two i's: with and without a dot, and it
# has a different rule for capitalising them:
str_to_upper(c("i", "ı"))

str_to_upper(c("i", "ı"), locale = "tr")
```

The locale is specified as a ISO 639 language code, which is a two or three letter abbreviation. If you leave the locale blank, it will use the current locale, as provided by your operating system. Click [here](https://en.wikipedia.org/wiki/List_of_ISO_639-1_codes) to see a list of ISO codes.

The base R `order()` and `sort()` functions sort strings using the current locale. If you want robust behaviour across different computers, you may want to use str_sort() and str_order() which take an additional locale argument:

```{r}
x <- c("apple", "eggplant", "banana")

str_sort(x, locale = "en")  # English

str_sort(x, locale = "haw") # Hawaiian
```


# Matching patterns with regular expressions

Regexps are a very terse language that allow you to describe patterns in strings.To learn regular expressions, we can use `str_view()` and `str_view_all()`.

## Basic matches

The simplest patterns match exact strings:

```{r}
x <- c("apple", "banana", "pear")

str_view(x, "an")
```

The next step up in complexity is `.`, which matches any character (except a newline):

```{r}
str_view(x, ".a.")
```


## Anchors

By default, regular expressions will match any part of a string. It’s often useful to _anchor_ the regular expression so that it matches from the start or end of the string.

+ `^` to match the start of the string.

+ `$` to match the end of the string.

```{r}
x <- c("apple", "banana", "pear")

str_view(x, "^a")
```


```{r}
str_view(x, "a$")
```


## Character Classes and Alternatives

 You’ve already seen what `.` can do. There are four other useful tools:
 
+ `\d`: matches any digit.

+ `\s`: matches any whitespace (e.g. space, tab, newline).

+ `[abc]`: matches a, b, or c.

+ `[^abc]`: matches anything except a, b, or c.

Remember, to create a regular expression containing `\d` or `\s`, you’ll need to escape the `\` for the string, so you’ll type "`\\d`" or "`\\s`".

A character class containing a single character is a nice alternative to backslash escapes when you want to include a single metacharacter in a regex. Many people find this more readable.

```{r}
# Look for a literal character that normally has special meaning in a regex

str_view(c("abc", "a.c", "a*c", "a c"), "a[.]c")
```

```{r}
str_view(c("abc", "a.c", "a*c", "a c"), ".[*]c")
```

```{r}
str_view(c("abc", "a.c", "a*c", "a c"), "a[ ]")
```

This works for most (but not all) regex metacharacters: `$` `.` `|` `?` `*` `+` `(` `)` `[` {. Unfortunately, a few characters have special meaning even inside a character class and must be handled with backslash escapes: `]` `\` `^` and `-`.


You can use _alternation_ to pick between one or more alternative patterns. For example, `abc|d..f` will match either ‘“abc”’, or "deaf". Note that the precedence for `|` is low, so that `abc|xyz` matches `abc` or `xyz` not `abcyz` or `abxyz.` Like with mathematical expressions, if precedence ever gets confusing, use parentheses to make it clear what you want:

```{r}
str_view(c("grey", "gray"), "gr(e|a)y")
```


## Repetition

The next step up in power involves controlling how many times a pattern matches:

+ `?`: 0 or 1

+ `+`: 1 or more

+ `*`: 0 or more


```{r}
x <- "1888 is the longest year in Roman numerals: MDCCCLXXXVIII"

str_view(x, "CC?")
```

```{r}
str_view(x, "CC+")
```

```{r}
str_view(x, 'C[LX]+')
```

Note that the precedence of these operators is high, so you can write: `colou?r` to match either American or British spellings. That means most uses will need parentheses, like `bana(na)+`.

You can also specify the number of matches precisely:

+ `{n}`: exactly n

+ `{n,}`: n or more

+ `{,m}`: at most m

+ `{n,m}`: between n and m

```{r}
str_view(x, "C{2}")
```

```{r}
str_view(x, "C{2,}")
```

```{r}
str_view(x, "C{2,3}")
```

By default these matches are “greedy”: they will match the longest string possible. You can make them “lazy”, matching the shortest string possible by putting a `?` after them. This is an advanced feature of regular expressions, but it’s useful to know that it exists:

```{r}
str_view(x, 'C{2,3}?')
```

```{r}
str_view(x, 'C[LX]+?')
```


## Grouping and backreferences

In addition to disambiguating complex expressions, parentheses also create a numbered capturing group (number 1, 2 etc.). A capturing group stores the part of the string matched by the part of the regular expression inside the parentheses. You can refer to the same text as previously matched by a capturing group with backreferences, like `\1`, `\2` etc. For example, the following regular expression finds all fruits that have a repeated pair of letters.

```{r}
str_view(fruit, "(..)\\1", match = TRUE)
```

# Tools

A word of caution before we continue: because regular expressions are so powerful, it’s easy to try and solve every problem with a single regular expression. In the words of Jamie Zawinski:

> Some people, when confronted with a problem, think “I know, I’ll use regular expressions.” Now they have two problems.


## Detect Matches

To determine if a character vector matches a pattern, `use str_detect()`.

```{r}
x <- c("apple", "banana", "pear")

str_detect(x, "e")
```

Remember that when you use a logical vector in a numeric context, `FALSE` becomes 0 and `TRUE` becomes 1. That makes `sum()` and `mean()` useful if you want to answer questions about matches across a larger vector:

```{r}
# How many common words start with t?
sum(str_detect(words, "^t"))

# What proportion of common words end with a vowel?
mean(str_detect(words, "[aeiou]$"))
```

When you have complex logical conditions (e.g. match a or b but not c unless d) it’s often easier to combine multiple `str_detect()` calls with logical operators, rather than trying to create a single regular expression. For example, here are two ways to find all words that don’t contain any vowels:

```{r}
# Find all words containing at least one vowel, and negate

no_vowels_1 <- !str_detect(words, "[aeiou]")

# Find all words consisting only of consonants (non-vowels)

no_vowels_2 <- str_detect(words, "^[^aeiou]+$")

identical(no_vowels_1, no_vowels_2)
```

A common use of `str_detect()` is to select the elements that match a pattern. You can do this with logical subsetting, or the convenient `str_subset()` wrapper:

```{r}
words[str_detect(words, "x$")]
```

```{r}
str_subset(words, "x$")
```

Typically, however, your strings will be one column of a data frame, and you’ll want to use `filter()` instead:

```{r}
df <- tibble(
  
  word = words, 
  
  i = seq_along(word)
  
)

df %>% 
  
  filter(str_detect(word, "x$"))
```


A variation on `str_detect()` is `str_count()`: rather than a simple yes or no, it tells you how many matches there are in a string:

```{r}
x <- c("apple", "banana", "pear")

str_count(x, "a")

# On average, how many vowels per word?

mean(str_count(words, "[aeiou]"))
```


It’s natural to use `str_count()` with `mutate()`:

```{r}
df %>% 
  
  mutate(
    
    vowels = str_count(word, "[aeiou]"),
    
    consonants = str_count(word, "[^aeiou]")
    
  )
```

Note that matches never overlap. For example, in `"abababa"`, how many times will the pattern `"aba"` match? Regular expressions say two, not three:


```{r}
str_count("abababa", "aba")

str_view_all("abababa", "aba")
```


## Extract Matches

To extract the actual text of a match, use `str_extract()`. Let's use [Harvard Sentences](https://en.wikipedia.org/wiki/Harvard_sentences), also found in `stringr::sentences` to practice on:

```{r}
length(sentences)

head(sentences)
```

Imagine we want to find all sentences that contain a colour. We first create a vector of colour names, and then turn it into a single regular expression:

```{r}
colours <- c("red", "orange", "yellow", "green", "blue", "purple")

colour_match <- str_c(colours, collapse = "|")

colour_match
```

Now we can select the sentences that contain a colour, and then extract the colour to figure out which one it is:

```{r}
has_colour <- str_subset(sentences, colour_match)

matches <- str_extract(has_colour, colour_match)

head(matches)
```

Note that `str_extract()` only extracts the first match. We can see that most easily by first selecting all the sentences that have more than 1 match:

```{r}
more <- sentences[str_count(sentences, colour_match) > 1]

str_view_all(more, colour_match)

str_extract(more, colour_match)
```

This is a common pattern for stringr functions, because working with a single match allows you to use much simpler data structures. To get all matches, use `str_extract_all()`. It returns a list:

```{r}
str_extract_all(more, colour_match)
```

If you use `simplify = TRUE`, `str_extract_all()` will return a matrix with short matches expanded to the same length as the longest:

```{r}
str_extract_all(more, colour_match, simplify = TRUE)


x <- c("a", "a b", "a b c")

str_extract_all(x, "[a-z]", simplify = TRUE)
```


## Grouped Matches