---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway:

```{r load}
if (!require("tidyverse")) install.packages("tidyverse")
```

The focus of this chapter will be on factors. In R, factors are used to work with categorical variables, variables that have a fixed and known set of possible values. They are also useful when you want to display character vectors in a non-alphabetical order.

# Creating Factors

Often, using character vectors messes with ordering and, if you're repeating vectors, it's easy to make typo's. To create a factor you must start by creating a list of the valid **levels**:

```{r}
month_levels <- c(
  
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
  
)

x1 <- c("Dec", "Apr", "Jan", "Mar")
x2 <- c("Dec", "Apr", "Jam", "Mar") # Meant to contain a typo
```

Now you can create a factor for a subset of months, for example:

```{r}
y1 <- factor(x1, levels = month_levels)
y1

sort(y1)
```

Any values not in the set will be silently converted to NA:

```{r}
y2 <- factor(x2, levels = month_levels)

y2
```

If you want a warning, you can use `readr::parse_factor()`:

```{r}
y2 <- parse_factor(x2, levels = month_levels)
```


If you omit the levels, they’ll be taken from the data in alphabetical order:

```{r}
factor(x1)
```

if you prefer that the order of the levels match the order of the first appearance in the data, use the following:

```{r}
f1 <- factor(x1, levels = unique(x1))

f1

# OR

f2 <- x1 %>% factor() %>% fct_inorder()

f2
```


If you ever need to access the set of valid levels directly, you can do so with `levels()`:

# General Social Survey

Moving forward we're going to use `forcats::gss_cat` as an example dataset. 

```{r}
forcats::gss_cat
```

When factors are stored in a tibble, you can’t see their levels so easily. One way to see them is with `count()`:

```{r}
gss_cat %>%
  
  count(race)
```

Also with a barchart:

```{r}
ggplot(gss_cat, aes(race)) +
  
  geom_bar()
```

By default, ggplot2 will drop levels that don’t have any values. You can force them to display with:

```{r}
ggplot(gss_cat, aes(race)) +

    geom_bar() +

    scale_x_discrete(drop = FALSE)
```


# Modifying Factor Order

It’s often useful to change the order of the factor levels in a visualisation. You can go from 

```{r}
relig_summary <- 
  gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(tvhours, relig)) + geom_point()
```

to the following

```{r}
ggplot(relig_summary, aes(tvhours, fct_reorder(relig, tvhours))) +
  
  geom_point()
```

by using `fct_reorder()`. It takes three arguments:

+ `f`, the factor whose levels you want to modify.

+ `x`, a numeric vector that you want to use to reorder the levels.

+ Optionally, `fun`, a function that’s used if there are multiple values of `x` for each value of `f`. The default value is `median`.


It is, however, better to reserve `fct_reorder()` for factors whose levels are arbitrarily ordered. Reordering according to income, for instance, doesn't make sense:

```{r}
rincome_summary <- gss_cat %>%
  
  group_by(rincome) %>%
  
  summarise(
  
      age = mean(age, na.rm = TRUE),
    
      tvhours = mean(tvhours, na.rm = TRUE),
    
      n = n()
  
    )

ggplot(rincome_summary, aes(age, fct_reorder(rincome, age))) + geom_point()
```


`fct_relevel()` takes the following arguments: 

  + factor, `f`, and then any number of levels that you want to move to the front of the line. Here we move "Not Applicable":

```{r}
ggplot(rincome_summary, aes(age, fct_relevel(rincome, "Not applicable"))) +
  
  geom_point()
```

Another type of reordering is useful when you are colouring the lines on a plot. `fct_reorder2()` reorders the factor by the `y` values associated with the largest `x` values. This makes the plot easier to read because the line colours line up with the legend.

```{r}
by_age <- gss_cat %>%
  
  filter(!is.na(age)) %>%
  
  count(age, marital) %>%
  
  group_by(age) %>%
  
  mutate(prop = n / sum(n))


ggplot(by_age, aes(age, prop, colour = marital)) +
  
  geom_line(na.rm = TRUE)


ggplot(by_age, aes(age, prop, colour = fct_reorder2(marital, age, prop))) +

  geom_line() +
  
  labs(colour = "marital")
```

Finally, for bar plots, you can use `fct_infreq()` to order levels in increasing frequency: this is the simplest type of reordering because it does not need any extra variables. You may want to combine with `fct_rev()`.

```{r}
gss_cat %>%
  
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>%
  
  ggplot(aes(marital)) +
  
  geom_bar()
```

# Modifying factor Levels

More powerful than changing the orders of the levels is changing their values. This allows you to clarify labels for publication, and collapse levels for high-level displays.The most general and powerful tool is `fct_recode()`. It allows you to recode, or change, the value of each level. For example, take the `gss_cat$partyid`:

```{r}
gss_cat %>% count(partyid)
```

The levels are terse and inconsistent

```{r}
gss_cat %>%
  
  mutate(partyid = fct_recode(partyid,
                              
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat"
    
  )) %>%
  
  count(partyid)
```

`fct_recode()` will leave levels that aren’t explicitly mentioned as is, and will warn you if you accidentally refer to a level that doesn’t exist.

To combine groups, you can assign multiple old levels to the same new level:

```{r}
gss_cat %>%
  
  mutate(partyid = fct_recode(partyid,
                              
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat",
    "Other"                 = "No answer",
    "Other"                 = "Don't know",
    "Other"                 = "Other party"
    
  )) %>%
  
  count(partyid)
```

If you want to collapse a lot of levels, `fct_collapse()` is a useful variant of `fct_recode()`. For each new variable, you can provide a vector of old levels:

```{r}
gss_cat %>%
  
  mutate(partyid = fct_collapse(partyid,
                                
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
    
  )) %>%
  
  count(partyid)
```

Sometimes you just want to lump together all the small groups to make a plot or table simpler. That’s the job of `fct_lump()`:

```{r}
gss_cat %>%
  
  mutate(relig = fct_lump(relig)) %>%
  
  count(relig)
```

The default behaviour is to progressively lump together the smallest groups, ensuring that the aggregate is still the smallest group. Instead, we can use the n parameter to specify how many groups (excluding other) we want to keep:

```{r}
gss_cat %>%
  
  mutate(relig = fct_lump(relig, n = 10)) %>%
  
  count(relig, sort = TRUE) %>%
  print(n = Inf)
```

