---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway. We Also load the `lubridate` package and the `nycflights13` dataset. 

```{r load}
if (!require("tidyverse")) install.packages("tidyverse")

library(lubridate)
library(nycflights13)
```

This chapter will show you how to work with dates and times in R.

# Creating date/times

There are three types of date/time data that refer to an instant in time:

1. A **date**. Tibbles print this as `<date>`.

2. A **time** within a day. Tibbles print this as `<time>`.

3. A **date-time** is a date plus a time: it uniquely identifies an instant in time (typically to the nearest second). Tibbles print this as `<dttm>`. Elsewhere in R these are called POSIXct, but I don’t think that’s a very useful name.

In this chapter we are only going to focus on dates and date-times as R doesn’t have a native class for storing times. If you need one, you can use the *hms* package.

To get the current date or date-time you can use `today()` or `now()`:

```{r}
today()

now()
```

You can also create a date/time from a string, from individual date-time components and from an existing date/time object.