---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway. We Also load the `lubridate` package and the `nycflights13` dataset. 

```{r load}
if (!require("tidyverse")) install.packages("tidyverse")

library(lubridate)
library(nycflights13)
```

This chapter will show you how to work with dates and times in R.

# Creating date/times

There are three types of date/time data that refer to an instant in time:

1. A **date**. Tibbles print this as `<date>`.

2. A **time** within a day. Tibbles print this as `<time>`.

3. A **date-time** is a date plus a time: it uniquely identifies an instant in time (typically to the nearest second). Tibbles print this as `<dttm>`. Elsewhere in R these are called POSIXct, but I don’t think that’s a very useful name.

In this chapter we are only going to focus on dates and date-times as R doesn’t have a native class for storing times. If you need one, you can use the *hms* package.

## From Nothing

To get the current date or date-time you can use `today()` or `now()`:

```{r}
today()

now()
```

## From Strings

You can also create a date/time from a string, from individual date-time components and from an existing date/time object.

Date/time data often comes as strings. One approach to parsing strings into date-times to use the helpers provided by lubridate. They automatically work out the format once you specify the order of the component. To use them, identify the order in which year, month, and day appear in your dates, then arrange “y”, “m”, and “d” in the same order.

```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")

```

These functions also take unquoted numbers. 

```{r}
ymd(20170131)
```
`ymd()` and friends create dates. To create a date-time, add an underscore and one or more of “h”, “m”, and “s” to the name of the parsing function:

```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")
```

You can also force the creation of a date-time from a date by supplying a timezone:

```{r}
ymd(20170131, tz = "UTC")
```

## From individual components

Instead of a single string, sometimes you’ll have the individual components of the date-time spread across multiple columns. This is what we have in the flights data:

```{r}
flights %>% 
  select(year, month, day, hour, minute)
```

To create a date/time from this sort of input, use `make_date()` for dates, or `make_datetime()` for date-times:

```{r}
flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))
```

Let’s do the same thing for each of the four time columns in `flights.` The times are represented in a slightly odd format, so we use modulus arithmetic to pull out the hour and minute components.

```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt
```

With this data, we can visualise the distribution of departure times across the year:

```{r}
flights_dt %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day
```

Or within a single day:

```{r}
flights_dt %>% 
  filter(dep_time < ymd(20130102)) %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 600) # 600 s = 10 minutes
```

Note that when you use date-times in a numeric context (like in a histogram), 1 means 1 second, so a binwidth of 86400 means one day. For dates, 1 means 1 day.

## From other types

You may want to switch between a date-time and a date. That’s the job of `as_datetime()` and `as_date()`:

```{r}
as_datetime(today())
as_date(now())
```

Sometimes you’ll get date/times as numeric offsets from the “Unix Epoch”, 1970-01-01. If the offset is in seconds, use `as_datetime()`; if it’s in days, `use as_date()`.

```{r}
as_datetime(60 * 60 * 10)
as_date(365 * 10 + 2)
```

# Date-time components

This section will focus on the accessor functions that let you get and set individual components. The next section will look at how arithmetic works with date-times.

## Getting components

You can pull out individual parts of the date with the accessor functions `year()`, `month()`, `mday()` (day of the month), `yday()` (day of the year), `wday()` (day of the week), `hour()`, `minute()`, and `second()`.

```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")

year(datetime)
month(datetime)
mday(datetime)

yday(datetime)
wday(datetime)
```

For `month()` and `wday()` you can set label = TRUE to return the abbreviated name of the month or day of the week. Set `abbr = FALSE` to return the full name.

```{r}
month(datetime, label = TRUE)

wday(datetime, label = TRUE, abbr = FALSE)
```

We can use `wday()` to see that more flights depart during the week than on the weekend:

```{r}
flights_dt %>% 
  mutate(wday = wday(dep_time, label = TRUE)) %>% 
  ggplot(aes(x = wday)) +
    geom_bar()
```

There’s an interesting pattern if we look at the average departure delay by minute within the hour. It looks like flights leaving in minutes 20-30 and 50-60 have much lower delays than the rest of the hour!

```{r}
flights_dt %>% 
  mutate(minute = minute(dep_time)) %>% 
  group_by(minute) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, avg_delay)) +
    geom_line()
```

Interestingly, if we look at the *scheduled* departure time we don’t see such a strong pattern:

```{r}
sched_dep <- flights_dt %>% 
  mutate(minute = minute(sched_dep_time)) %>% 
  group_by(minute) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n())

ggplot(sched_dep, aes(minute, avg_delay)) +
  geom_line()
```

So why do we see that pattern with the actual departure times? Well, like much data collected by humans, there’s a strong bias towards flights leaving at “nice” departure times.

```{r}
ggplot(sched_dep, aes(minute, n)) +
  geom_line()
```

## Rounding

An alternative approach to plotting individual components is to round the date to a nearby unit of time, with `floor_date()`, `round_date()`, and `ceiling_date()`. Each function takes a vector of dates to adjust and then the name of the unit round down (floor), round up (ceiling), or round to. This, for example, allows us to plot the number of flights per week:

```{r}
flights_dt %>% 
  count(week = floor_date(dep_time, "week")) %>% 
  ggplot(aes(week, n)) +
    geom_line()
```

Computing the difference between a rounded and unrounded date can be particularly useful.


## Setting components

You can also use each accessor function to set the components of a date/time:

```{r}
(datetime <- ymd_hms("2016-07-08 12:34:56"))

year(datetime) <- 2020
datetime

month(datetime) <- 01
datetime

hour(datetime) <- hour(datetime) + 1
datetime
```

Alternatively, rather than modifying in place, you can create a new date-time with `update()`. This also allows you to set multiple values at once.

```{r}
update(datetime, year = 2020, month = 2, mday = 2, hour = 2)
```

You can use `update()` to show the distribution of flights across the course of the day for every day of the year:

```{r}
flights_dt %>% 
  mutate(dep_hour = update(dep_time, yday = 1)) %>% 
  ggplot(aes(dep_hour)) +
    geom_freqpoly(binwidth = 300)
```

Setting larger components of a date to a constant is a powerful technique that allows you to explore patterns in the smaller components.

## Durations

In R, when you subtract two dates, you get a difftime object:

```{r}
# How old is Hadley?
h_age <- today() - ymd(19791014)
h_age
```

A difftime class object records a time span of seconds, minutes, hours, days, or weeks. This ambiguity can make difftimes a little painful to work with, so lubridate provides an alternative which always uses seconds: the **duration**.

```{r}
as.duration(h_age)
```

```{r}
dseconds(15)
dminutes(10)
dhours(c(12, 24))
ddays(0:5)
dweeks(3)
dyears(1)
```

You can add and multiply durations:

```{r}
2 * dyears(1)
dyears(1) + dweeks(12) + dhours(15)
```


You can add and subtract durations to and from days:

```{r}
tomorrow <- today() + ddays(1)
last_year <- today() - dyears(1)
```

However, because durations represent an *exact* number of seconds, sometimes you might get an unexpected result:

```{r}
one_pm <- ymd_hms("2016-03-12 13:00:00", tz = "America/New_York")

one_pm
one_pm + ddays(1)
```

Why is one day after 1pm on March 12, 2pm on March 13?! If you look carefully at the date you might also notice that the time zones have changed. Because of DST, March 12 only has 23 hours, so if we add a full days worth of seconds we end up with a different time.

## Periods

To solve this problem, lubridate provides periods. Periods are time spans but don’t have a fixed length in seconds, instead they work with “human” times, like days and months. That allows them to work in a more intuitive way:

```{r}
one_pm
one_pm + days(1)
```

Like durations, periods can be created with a number of friendly constructor functions.

```{r}
seconds(15)
minutes(10)
hours(c(12, 24))
days(7)
months(1:6)
weeks(3)
years(1)
```

You can add and multiply periods:

```{r}
10 * (months(6) + days(1))
days(50) + hours(25) + minutes(2)
```

And of course, add them to dates. Compared to durations, periods are more likely to do what you expect:

```{r}
# A leap year
ymd("2016-01-01") + dyears(1)
ymd("2016-01-01") + years(1)

# Daylight Savings Time
one_pm + ddays(1)
one_pm + days(1)
```

Let’s use periods to fix an oddity related to our flight dates. Some planes appear to have arrived at their destination before they departed from New York City.

```{r}
flights_dt %>% 
  filter(arr_time < dep_time) 
```

These are overnight flights. We used the same date information for both the departure and the arrival times, but these flights arrived on the following day. We can fix this by adding days(1) to the arrival time of each overnight flight.

```{r}
flights_dt <- flights_dt %>% 
  mutate(
    overnight = arr_time < dep_time,
    arr_time = arr_time + days(overnight * 1),
    sched_arr_time = sched_arr_time + days(overnight * 1)
  )
```

Now all of our flights obey the laws of physics.

```{r}
flights_dt %>% 
  filter(overnight, arr_time < dep_time) 
```


## Intevals

It’s obvious what dyears(1) / ddays(365) should return: one, because durations are always represented by a number of seconds, and a duration of a year is defined as 365 days worth of seconds.

```{r}
years(1) / days(1)
```

If you want a more accurate measurement, you’ll have to use an **interval**. An interval is a duration with a starting point: that makes it precise so you can determine exactly how long it is:

```{r}
next_year <- today() + years(1)
(today() %--% next_year) / ddays(1)
```

To find out how many periods fall into an interval, you need to use integer division:

```{r}
(today() %--% next_year) %/% days(1)
```

## Conclusion

How do you pick between duration, periods, and intervals? As always, pick the simplest data structure that solves your problem. If you only care about physical time, use a duration; if you need to add human times, use a period; if you need to figure out how long a span is in human units, use an interval.

To see a summary of the allowed arithmetic between different data types, see (this summary)[https://r4ds.had.co.nz/dates-and-times.html#summary-1].

# Time Zones

Time zones are an enormously complicated topic because of their interaction with geopolitical entities. Fortunately we don’t need to dig into all the details as they’re not all important for data analysis, but there are a few challenges we’ll need to tackle head on.

The first challenge is that everyday names of time zones tend to be ambiguous. For example, if you’re American you’re probably familiar with EST, or Eastern Standard Time. However, both Australia and Canada also have EST! To avoid confusion, R uses the international standard IANA time zones. These use a consistent naming scheme “/”, typically in the form “<continent>/<city>” (there are a few exceptions because not every country lies on a continent). Examples include “America/New_York”, “Europe/Paris”, and “Pacific/Auckland”.

You might wonder why the time zone uses a city, when typically you think of time zones as associated with a country or region within a country. This is because the IANA database has to record decades worth of time zone rules. In the course of decades, countries change names (or break apart) fairly frequently, but city names tend to stay the same. Another problem is that name needs to reflect not only to the current behaviour, but also the complete history. For example, there are time zones for both “America/New_York” and “America/Detroit”. These cities both currently use Eastern Standard Time but in 1969-1972 Michigan (the state in which Detroit is located), did not follow DST, so it needs a different name. It’s worth reading the raw time zone database (available at http://www.iana.org/time-zones) just to read some of these stories!

You can find out what R thinks your current time zone is with `Sys.timezone()`:

```{r}
Sys.timezone()
```

(If R doesn’t know, you’ll get an NA.)

And see the complete list of all time zone names with `OlsonNames()`:

```{r}
length(OlsonNames())
head(OlsonNames())
```

In R, the time zone is an attribute of the date-time that only controls printing. For example, these three objects represent the same instant in time:

```{r}
(x1 <- ymd_hms("2015-06-01 12:00:00", tz = "America/New_York"))
(x2 <- ymd_hms("2015-06-01 18:00:00", tz = "Europe/Copenhagen"))
(x3 <- ymd_hms("2015-06-02 04:00:00", tz = "Pacific/Auckland"))

x1 - x2
x1 - x3
```

Unless otherwise specified, lubridate always uses UTC. UTC (Coordinated Universal Time) is the standard time zone used by the scientific community and roughly equivalent to its predecessor GMT (Greenwich Mean Time). It does not have DST, which makes a convenient representation for computation. Operations that combine date-times, like c(), will often drop the time zone. In that case, the date-times will display in your local time zone:

```{r}
x4 <- c(x1, x2, x3)
x4
```

You can change the time zone in two ways:

+ Keep the instant in time the same, and change how it’s displayed. Use this when the instant is correct, but you want a more natural display.

```{r}
x4a <- with_tz(x4, tzone = "Australia/Lord_Howe")
x4a

x4a - x4
```

(This also illustrates another challenge of times zones: they’re not all integer hour offsets!)

+ Change the underlying instant in time. Use this when you have an instant that has been labelled with the incorrect time zone, and you need to fix it.

```{r}
x4b <- force_tz(x4, tzone = "Australia/Lord_Howe")
x4b

x4b - x4
```



