---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway. We Also load the `lubridate` package and the `nycflights13` dataset. 

```{r load}
if (!require("tidyverse")) install.packages("tidyverse")

library(lubridate)
library(nycflights13)
```

This chapter will show you how to work with dates and times in R.

# Creating date/times

There are three types of date/time data that refer to an instant in time:

1. A **date**. Tibbles print this as `<date>`.

2. A **time** within a day. Tibbles print this as `<time>`.

3. A **date-time** is a date plus a time: it uniquely identifies an instant in time (typically to the nearest second). Tibbles print this as `<dttm>`. Elsewhere in R these are called POSIXct, but I don’t think that’s a very useful name.

In this chapter we are only going to focus on dates and date-times as R doesn’t have a native class for storing times. If you need one, you can use the *hms* package.

## From Nothing

To get the current date or date-time you can use `today()` or `now()`:

```{r}
today()

now()
```

## From Strings

You can also create a date/time from a string, from individual date-time components and from an existing date/time object.

Date/time data often comes as strings. One approach to parsing strings into date-times to use the helpers provided by lubridate. They automatically work out the format once you specify the order of the component. To use them, identify the order in which year, month, and day appear in your dates, then arrange “y”, “m”, and “d” in the same order.

```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")

```

These functions also take unquoted numbers. 

```{r}
ymd(20170131)
```
`ymd()` and friends create dates. To create a date-time, add an underscore and one or more of “h”, “m”, and “s” to the name of the parsing function:

```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")
```

You can also force the creation of a date-time from a date by supplying a timezone:

```{r}
ymd(20170131, tz = "UTC")
```

## From individual components

Instead of a single string, sometimes you’ll have the individual components of the date-time spread across multiple columns. This is what we have in the flights data:

```{r}
flights %>% 
  select(year, month, day, hour, minute)
```

To create a date/time from this sort of input, use `make_date()` for dates, or `make_datetime()` for date-times:

```{r}
flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))
```

