---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

First, we install/load the relevant packages by installing the [tidyverse](https://www.tidyverse.org/) library. This "package" effectively contains all packages developed by *the* [Hadley Wickham](http://hadley.nz/) that relate to the [tidy data](https://www.jstatsoft.org/article/view/v059i10) way of thinking. This chapter focuses on specific libraries, but we install and load the whole `tidyverse` anyway:

```{r setup}
if (!require("tidyverse")) install.packages("tidyverse")
```

# Pipes

The pipe, `%>%`, comes from the magrittr package by Stefan Milton Bache. Packages in the tidyverse load %>% for you automatically, so you don’t usually load **magrittr** explicitly. Here, however, we’re focussing on piping, and we aren’t loading any other packages, so we will load it explicitly.

```{r}
library(magrittr)
```

## Piping alternatives

The point of the pipe is to help you write code in a way that is easier to read and understand.The point of the pipe is to help you write code in a way that is easier to read and understand. Without the pipe, you might write code that needs to store the result at each step. If there are natural names, this is a good idea, and you should do it. But many times there aren’t natural names, and you add numeric suffixes to make the names unique. That leads to two problems:

1. The code is cluttered with unimportant names
2. You have to carefully increment the suffix on each line.

Memory should generally not be a problem, though. R isn’t stupid, and it will share columns across data frames, where possible. Let’s take a look at an actual data manipulation pipeline where we add a new column to `ggplot2::diamonds`: 

```{r}
diamonds <- ggplot2::diamonds
diamonds2 <- diamonds %>% 
  dplyr::mutate(price_per_carat = price / carat)

pryr::object_size(diamonds)
pryr::object_size(diamonds2)
pryr::object_size(diamonds, diamonds2)
```

From above, you can see that `diamonds2` has 10 columns in common with `diamonds`: there’s no need to duplicate all that data, so the two data frames have| variables in common. These variables will only get copied if you modify one of them. 

### Overwrite the original

Instead of creating intermediate objects at each step, we could overwrite the original object.This is less typing (and less thinking), so you’re less likely to make mistakes. However, there are two problems:

1. Debugging is painful: if you make a mistake you’ll need to re-run the complete pipeline from the beginning.
2. The repetition of the object being transformed (we’ve written foo_foo six times!) obscures what’s changing on each line.

### Function composition

Another approach is to abandon assignment and just string the function calls together:
```{r}
bop(
  scoop(
    hop(foo_foo, through = forest),
    up = field_mice
  ), 
  on = head
)
```

Here the disadvantage is that you have to read from inside-out, from right-to-left, and that the arguments end up spread far apart, making it more difficult to read.

## We Could Just use the Pipe

The pipe focuses on verbs, not nouns. The pipe enables you to read a series of function compositions like it’s a set of imperative actions. This is useful for understanding the flow of code. The pipe (`%>%`) works by performing a “lexical transformation”: behind the scenes, magrittr reassembles the code in the pipe to a form that works by overwriting an intermediate object. When you run a pipe, magrittr does something like this:

```{r}
my_pipe <- function(.) {
  . <- hop(., through = forest)
  . <- scoop(., up = field_mice)
  bop(., on = head)
}
my_pipe(foo_foo)
```

This means that the pipe won’t work for two classes of functions:

1. Functions that use the current environment. For example, assign() will create a new variable with the given name in the current environment:
```{r}
assign("x", 10)
"x" %>% assign(100)
```

The use of assign with the pipe does not work because it assigns it to a temporary environment used by %>%. If you do want to use assign with the pipe, you must be explicit about the environment:
```{r}
env <- environment()
"x" %>% assign(100, envir = env)
x
```


2. Functions that use lazy evaluation. In R, function arguments are only computed when the function uses them, not prior to calling the function. The pipe computes each element in turn, so you can’t rely on this behaviour.

One place that this is a problem is `tryCatch()`, which lets you capture and handle errors:

```{r}
tryCatch(stop("!"), error = function(e) "An error")

stop("!") %>% 
  tryCatch(error = function(e) "An error")
```

There are a relatively wide class of functions with this behaviour, including try(), suppressMessages(), and suppressWarnings() in base R.

## When not to use the pipe

Pipes are most useful for rewriting a fairly short linear sequence of operations.Reach for another tool when:

+ Your pipes are longer than (say) ten steps. In that case, create intermediate objects with meaningful names. That will make debugging easier, because you can more easily check the intermediate results, and it makes it easier to understand your code, because the variable names can help communicate intent.

+ You have multiple inputs or outputs. If there isn’t one primary object being transformed, but two or more objects being combined together, don’t use the pipe.

+ You are starting to think about a directed graph with a complex dependency structure. Pipes are fundamentally linear and expressing complex relationships with them will typically yield confusing code.


## Other tools from magrittr 

There are some other useful tools inside magrittr that you might want to try out:

+ When working with more complex pipes, it’s sometimes useful to call a function for its side-effects. Maybe you want to print out the current object, or plot it, or save it to disk. Many times, such functions don’t return anything, effectively terminating the pipe. To work around this problem, you can use the “tee” pipe. [%T>%](https://magrittr.tidyverse.org/reference/tee.html) works like %>% except that it returns the left-hand side instead of the right-hand side. It’s called “tee” because it’s like a literal T-shaped pipe.

```{r}
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()

rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
```

+ If you’re working with functions that don’t have a data frame based API (i.e. you pass them individual vectors, not a data frame and expressions to be evaluated in the context of that data frame), you might find [%$%](https://magrittr.tidyverse.org/reference/exposition.html) useful. It “explodes” out the variables in a data frame so that you can refer to them explicitly. This is useful when working with many functions in base R:

+ For assignment magrittr provides the %<>% operator which allows you to replace code like:

```{r}
mtcars <- mtcars %>% 
  transform(cyl = cyl * 2)
```

with

```{r}
mtcars %<>% transform(cyl = cyl * 2)
```

