---
output:
  html_document:
    toc: true
    toc_float: true
    theme: lumen
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

<br>

This section is part of a series that serves as a condensed summary of the [R for Data Science](http://r4ds.had.co.nz/index.html) book as I work through it. The purpose is to use this series as a quick reference in future.

The focus of this chapter is on writing functions in base R, so you won’t need any extra packages.

# Vectors

Vectors are particularly important as most of the functions you will write will work with vectors. The focus of this chapter is on base R data structures, so it isn’t essential to load any packages. We will, however, use a handful of functions from the purrr package to avoid some inconsistencies in base R.

```{r}
library(tidyverse)
```

## Vector Basics

There are two types of vectors:

1. *Atomic* vectors, of which there are six types: *logical*, *integer*, *double*, *character*, *complex*, and *raw*. Integer and double vectors are collectively known as numeric vectors.

2. *Lists*, which are sometimes called recursive vectors because lists can contain other lists.

The chief difference between atomic vectors and lists is that atomic vectors are *homogeneous*, while lists can be *heterogeneous*. There’s one other related object: `NULL`. `NULL` is often used to represent the absence of a vector (as opposed to `NA` which is used to represent the absence of a value in a vector). `NULL` typically behaves like a vector of length 0. 

Every vector has two key properties:

1. Its *type*, which you can determine with `typeof()`.

```{r}
typeof(letters)
typeof(1:10)
```

2. Its *length*, which you can determine with `length()`.

```{r}
x <- list("a", "b", 1:10)
length(x)
```

Vectors can also contain arbitrary additional metadata in the form of attributes. These attributes are used to create *augmented vectors* which build on additional behaviour. There are three important types of augmented vector:

- Factors are built on top of integer vectors.

- Dates and date-times are built on top of numeric vectors.

- Data frames and tibbles are built on top of lists.

## Important types of atomic vector

The four most important types of atomic vector are logical, integer, double, and character.

### Logical

Logical vectors are the simplest type of atomic vector because they can take only three possible values: `FALSE`, `TRUE`, and `NA`. Logical vectors are usually constructed with comparison operators.

### Numeric

Integer and double vectors are known collectively as numeric vectors. In R, numbers are doubles by default. To make an integer, place an `L` after the number:

```{r}
typeof(1)
typeof(1L)
1.5L
```

The distinction between integers and doubles is not usually important, but there are two important differences that you should be aware of:

1. Doubles are approximations. Doubles represent floating point numbers that can not always be precisely represented with a fixed amount of memory. This means that you should consider all doubles to be approximations. For example, what is square of the square root of two?

```{r}
x <- sqrt(2) ^ 2
x
x - 2
```

This behaviour is common when working with floating point numbers: most calculations include some approximation error. Instead of comparing floating point numbers using `==`, you should use `dplyr::near()` which allows for some numerical tolerance.

2. Integers have one special value: `NA`, while doubles have four: `NA`, `NaN`, `Inf` and `-Inf`. All three special values `NaN`, `Inf` and `-Inf` can arise during division:

```{r}
c(-1, 0, 1) / 0
```

Avoid using `==` to check for these other special values. Instead use the helper functions `is.finite()`, `is.infinite()`, and `is.nan()`.

### Character

Character vectors are the most complex type of atomic vector, because each element of a character vector is a string, and a string can contain an arbitrary amount of data.

One important feature of the underlying string implementation: R uses a global string pool. This means that each unique string is only stored in memory once, and every use of the string points to that representation. This reduces the amount of memory needed by duplicated strings. You can see this behaviour in practice with `pryr::object_size()`:

```{r}
x <- "This is a reasonably long string."
pryr::object_size(x)

y <- rep(x, 1000)
pryr::object_size(y)
```

`y` doesn’t take up 1,000x as much memory as `x`, because each element of `y` is just a pointer to that same string. A pointer is 8 bytes, so 1000 pointers to a 152 B string is 8 * 1000 + 152 = 8.14 kB.

### Missing Values

Note that each type of atomic vector has its own missing value:

```{r}
NA            # logical
NA_integer_   # integer
NA_real_      # double
NA_character_ # character
```

Normally you don’t need to know about these different types because you can always use `NA` and it will be converted to the correct type using the implicit coercion rules described next. However, there are some functions that are strict about their inputs, so it’s useful to have this knowledge sitting in your back pocket so you can be specific when needed.

## USing Atomic Vectors

Now that you understand the different types of atomic vector, it’s useful to review some of the important tools for working with them. These include:

1. How to convert from one type to another, and when that happens automatically.

2. How to tell if an object is a specific type of vector.

3. What happens when you work with vectors of different lengths.

4. How to name the elements of a vector.

5. How to pull out elements of interest.

### Coercion

There are two ways to convert, or coerce, one type of vector to another:

1. Explicit coercion happens when you call a function like `as.logical()`, `as.integer()`, `as.double()`, or `as.character()`. Whenever you find yourself using explicit coercion, you should always check whether you can make the fix upstream, so that the vector never had the wrong type in the first place. For example, you may need to tweak your readr col_types specification.

2. Implicit coercion happens when you use a vector in a specific context that expects a certain type of vector. For example, when you use a logical vector with a numeric summary function, or when you use a double vector where an integer vector is expected.

We focus on implicit coercion. The most important type of implicit coercion uses a logical vector in a numeric context. In this case `TRUE` is converted to `1` and FALSE converted to `0`. That means the sum of a logical vector is the number of trues, and the mean of a logical vector is the proportion of trues:

```{r}
x <- sample(20, 100, replace = TRUE)
y <- x > 10
sum(y)  # how many are greater than 10?
mean(y) # what proportion are greater than 10?
```

It’s also important to understand what happens when you try and create a vector containing multiple types with `c()`: the most complex type always wins.

```{r}
typeof(c(TRUE, 1L))
typeof(c(1L, 1.5))
typeof(c(1.5, "a"))
```

An atomic vector can not have a mix of different types because the type is a property of the complete vector, not the individual elements. If you need to mix multiple types in the same vector, you should use a list.

### Test functions

Sometimes you want to do different things based on the type of vector. One option is to `use typeof()`. Another is to use a test function which returns a TRUE or FALSE. Base R provides many functions like `is.vector()` and `is.atomic()`, but they often return surprising results. Instead, it’s safer to use the `is_*` functions provided by purrr as summarised by this [table](https://r4ds.had.co.nz/vectors.html#test-functions).

### Scalars and recycling rules

As well as implicitly coercing the types of vectors to be compatible, R will also implicitly coerce the length of vectors. This is called vector *recycling*, because the shorter vector is repeated, or recycled, to the same length as the longer vector.

This is generally most useful when you are mixing vectors and “scalars”. I put scalars in quotes because R doesn’t actually have scalars: instead, a single number is a vector of length 1. Because there are no scalars, most built-in functions are vectorised, meaning that they will operate on a vector of numbers. That’s why, for example, this code works:

```{r}
sample(10) + 100

runif(10) > 0.5
```

In R, basic mathematical operations work with vectors. That means that you should never need to perform explicit iteration when performing simple mathematical computations.

It’s intuitive what should happen if you add two vectors of the same length, or a vector and a “scalar”, but what happens if you add two vectors of different lengths?

```{r}
1:10 + 1:2
```

Here, R will expand the shortest vector to the same length as the longest, so called recycling. This is silent except when the length of the longer is not an integer multiple of the length of the shorter:

```{r}
1:10 + 1:3
```

While vector recycling can be used to create very succinct, clever code, it can also silently conceal problems. For this reason, the vectorised functions in tidyverse will throw errors when you recycle anything other than a scalar. If you do want to recycle, you’ll need to do it yourself with `rep()`:

```{r eval=FALSE}
tibble(x = 1:4, y = 1:2)
tibble(x = 1:4, y = rep(1:2, 2))
tibble(x = 1:4, y = rep(1:2, each = 2))
```

### Naming vectors

All types of vectors can be named. You can name them during creation with `c()`:

```{r}
c(x = 1, y = 2, z = 4)
```

Or after the fact with `purrr::set_names()`:

```{r}
set_names(1:3, c("a", "b", "c"))
```

Named vectors are most useful for subsetting.

### Subsetting

So far we’ve used `dplyr::filter()` to filter the rows in a tibble. `filter()` only works with tibble, so we’ll need new tool for vectors: `[`. `[` is the subsetting function, and is called like `x[a]`. There are four types of things that you can subset a vector with:

1. A numeric vector containing only integers. The integers must either be all positive, all negative, or zero.

Subsetting with positive integers keeps the elements at those positions:

```{r}
x <- c("one", "two", "three", "four", "five")
x[c(3, 2, 5)]
```

By repeating a position, you can actually make a longer output than input:

```{r}
x[c(1, 1, 5, 5, 5, 2)]
```

Negative values drop the elements at the specified positions:

```{r eval=FALSE}
x[c(1, -1)]
```

The error message above mentions subsetting with zero, which returns no values:

```{r}
x[0]
```

This is not useful very often, but it can be helpful if you want to create unusual data structures to test your functions with.

2. Subsetting with a logical vector keeps all values corresponding to a `TRUE` value. This is most often useful in conjunction with the comparison functions.

```{r}
x <- c(10, 3, NA, 5, 8, 1, NA)
x[!is.na(x)]
x[x %% 2 == 0]
```

3. If you have a named vector, you can subset it with a character vector:

```{r}
x <- c(abc = 1, def = 2, xyz = 5)
x[c("xyz", "def")]
```

Like with positive integers, you can also use a character vector to duplicate individual entries.

4. The simplest type of subsetting is nothing, `x[]`, which returns the complete `x`. This is not useful for subsetting vectors, but it is useful when subsetting matrices (and other high dimensional structures) because it lets you select all the rows or all the columns, by leaving that index blank. For example, if `x` is 2d, `x[1, ]` selects the first row and all the columns, and `x[, -1]` selects all rows and all columns except the first.

There is an important variation of `[` called `[[`. `[[` only ever extracts a single element, and always drops names. It’s a good idea to use it whenever you want to make it clear that you’re extracting a single item, as in a for loop. The distinction between `[` and `[[` is most important for lists, as we’ll see shortly.

## Recursive vectors (lists)

Lists are a step up in complexity from atomic vectors, because lists can contain other lists. This makes them suitable for representing hierarchical or tree-like structures. You create a list with `list()`:

```{r}
x <- list(1, 2, 3)
x
```

A very useful tool for working with lists is `str()` because it focusses on the *structure*, not the contents.

```{r}
str(x)

x_named <- list(a = 1, b = 2, c = 3)
str(x_named)
```

Unlike atomic vectors, `list()` can contain a mix of objects:

```{r}
y <- list("a", 1L, 1.5, TRUE)
str(y)
```

Lists can even contain other lists!

```{r}
z <- list(list(1, 2), list(3, 4))
str(z)
```

### Visualising lists

view the digram [here](https://r4ds.had.co.nz/vectors.html#visualising-lists).

There are three principles:

1. Lists have rounded corners. Atomic vectors have square corners.

2. Children are drawn inside their parent, and have a slightly darker background to make it easier to see the hierarchy.

3. The orientation of the children (i.e. rows or columns) isn’t important, so I’ll pick a row or column orientation to either save space or illustrate an important property in the example.

### Subsetting

There are three ways to subset a list:

```{r}
a <- list(a = 1:3, b = "a string", c = pi, d = list(-1, -5))
```

+ `[` extracts a sub-list. The result will always be a list.

```{r}
str(a[1:2])
str(a[4])
```

Like with vectors, you can subset with a logical, integer, or character vector.

+ `[[` extracts a single component from a list. It removes a level of hierarchy from the list.

```{r}
str(a[[1]])
str(a[[4]])
```

+ `$` is a shorthand for extracting named elements of a list. It works similarly to `[[` except that you don’t need to use quotes.

```{r}
a$a
a[["a"]]
```

The distinction between `[` and `[[` is really important for lists, because `[[` drills down into the list while `[` returns a new, smaller list. Compare the code and output above with this [visual representation](https://r4ds.had.co.nz/vectors.html#visualising-lists).

## Attributes

Any vector can contain arbitrary additional metadata through its attributes. You can think of attributes as named list of vectors that can be attached to any object. You can get and set individual attribute values with `attr()` or see them all at once with `attributes()`.

```{r}
x <- 1:10
attr(x, "greeting")

attr(x, "greeting") <- "Hi!"
attr(x, "farewell") <- "Bye!"
attributes(x)
```

There are three very important attributes that are used to implement fundamental parts of R:

1. *Names* are used to name the elements of a vector.
2. *Dimensions* (dims, for short) make a vector behave like a matrix or array.
3. *Class* is used to implement the S3 object oriented system.

Something quick on class: Class controls how generic functions work. Generic functions are key to object oriented programming in R, because they make functions behave differently for different classes of input. Here’s what a typical generic function looks like:

```{r}
as.Date
```

The call to “UseMethod” means that this is a generic function, and it will call a specific *method*, a function, based on the class of the first argument. (All methods are functions; not all functions are methods). You can list all the methods for a generic with `methods()`:

```{r}
methods("as.Date")
```

For example, if `x` is a character vector, `as.Date()` will call `as.Date.character()`; if it’s a factor, it’ll call `as.Date.factor()`.

You can see the specific implementation of a method with `getS3method()`:

```{r}
getS3method("as.Date", "default")

getS3method("as.Date", "numeric")
```

The most important S3 generic is `print()`: it controls how the object is printed when you type its name at the console. Other important generics are the subsetting functions `[`, `[[`, and `$`.

## Augmented vectors

Atomic vectors and lists are the building blocks for other important vector types like factors and dates. I call these augmented vectors, because they are vectors with additional attributes, including class. Because augmented vectors have a class, they behave differently to the atomic vector on which they are built. In this book, we make use of four important augmented vectors:

+Factors

+Dates

+Date-times

+Tibbles

### Factors

Factors are designed to represent categorical data that can take a fixed set of possible values. Factors are built on top of integers, and have a levels attribute:

```{r}
x <- factor(c("ab", "cd", "ab"), levels = c("ab", "cd", "ef"))
typeof(x)
attributes(x)
```

### Dates and date-times

Dates in R are numeric vectors that represent the number of days since 1 January 1970.

```{r}
x <- as.Date("1971-01-01")
unclass(x)

typeof(x)

attributes(x)
```

Date-times are numeric vectors with class POSIXct that represent the number of seconds since 1 January 1970. (In case you were wondering, “POSIXct” stands for “Portable Operating System Interface”, calendar time.)

```{r}
x <- lubridate::ymd_hm("1970-01-01 01:00")
unclass(x)

typeof(x)
attributes(x)
```

The `tzone` attribute is optional. It controls how the time is printed, not what absolute time it refers to.

```{r}
attr(x, "tzone") <- "US/Pacific"
attr(x, "tzone") <- "US/Eastern"
```

There is another type of date-times called POSIXlt. These are built on top of named lists:

```{r}
y <- as.POSIXlt(x)
typeof(y)
attributes(y)
```

POSIXlts are rare inside the tidyverse. They do crop up in base R, because they are needed to extract specific components of a date, like the year or month. Since lubridate provides helpers for you to do this instead, you don’t need them. POSIXct’s are always easier to work with, so if you find you have a POSIXlt, you should always convert it to a regular data time `lubridate::as_date_time()`.

### Tibbles

Tibbles are augmented lists: they have class “tbl_df” + “tbl” + “data.frame”, and `names` (column) and `row.names` attributes:

```{r}
tb <- tibble::tibble(x = 1:5, y = 5:1)
typeof(tb)
attributes(tb)
```

The difference between a tibble and a list is that all the elements of a data frame must be vectors with the same length. All functions that work with tibbles enforce this constraint.

Traditional data.frames have a very similar structure:

```{r}
df <- data.frame(x = 1:5, y = 5:1)
typeof(df)
attributes(df)
```

The main difference is the class. The class of tibble includes “data.frame” which means tibbles inherit the regular data frame behaviour by default.


